---
title: "Final Project"
author: "Savanna Hatch"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup and data, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

asthma_data <- data.frame(
  state = c("Alabama","Alaska","Arizona","Arkansas","California","Colorado","Connecticut",
            "Delaware","District of Columbia","Florida","Georgia","Hawaii","Idaho","Illinois",
            "Indiana","Iowa","Kansas","Kentucky","Louisiana","Maine","Maryland","Massachusetts",
            "Michigan","Minnesota","Mississippi","Missouri","Montana","Nebraska","Nevada",
            "New Hampshire","New Jersey","New Mexico","New York","North Carolina","North Dakota",
            "Ohio","Oklahoma","Oregon","Pennsylvania","Rhode Island","South Carolina",
            "South Dakota","Tennessee","Texas","Utah","Vermont","Virginia","Washington",
            "West Virginia","Wisconsin","Wyoming"),
  
  state_abrv = c("AL","AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC", "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY"),
  
  number = c(378702,58742,555997,243032,2629546,500090,356651,
             78980,60410,1646583,802783,103228,161970,856488,
             574447,238689,237705,377365,351940,146031,499134,633126,
             935666,434473,211055,497216,102617,121412,251193,
             148330,644536,171212,1611027,764909,61944,
             1042354,373879,389831,1033741,117252,368883,
             56630,636322,1760654,266105,68112,671051,667228,
             181395,503318,46973),
  
  prevalence_raw = c("9.6 (0.59)","10.7 (0.58)","9.7 (0.49)","10.5 (0.57)","8.7 (0.39)","10.8 (0.43)","12.4 (0.51)",
                     "9.9 (0.64)","11.1 (0.77)","9.3 (0.53)","9.6 (0.48)","9.1 (0.45)","11.1 (0.51)","8.7 (0.58)",
                     "11.0 (0.39)","9.7 (0.42)","10.7 (0.41)","10.8 (0.68)","10.0 (0.55)","13.1 (0.52)","10.4 (0.38)","11.3 (0.43)",
                     "11.9 (0.44)","9.9 (0.32)","9.4 (0.63)","10.4 (0.49)","11.7 (0.51)","8.1 (0.46)","10.1 (0.79)",
                     "13.1 (0.64)","8.9 (0.43)","10.4 (0.62)","10.3 (0.34)","9.2 (0.62)","10.4 (0.60)",
                     "11.4 (0.38)","12.3 (0.54)","11.5 (0.51)","10.1 (0.67)","13.3 (0.64)","9.0 (0.40)",
                     "8.3 (0.75)","11.7 (0.61)","7.9 (0.37)","11.0 (0.43)","12.9 (0.53)","9.9 (0.46)","10.9 (0.26)",
                     "12.9 (0.61)","10.9 (0.42)","10.5 (0.65)"), 
  
  population = c(5108468, 733406, 7431344, 3067732, 38965193, 5877610, 3617176, 1031890, 678972,
  22610726, 11029227, 1435138, 1964726, 12549689, 6862199, 3207004, 2940546,
  4526154, 4573749, 1395722, 6180253, 7001399, 10037261, 5737915, 2939690,
  6196156, 1132812, 1978379, 3194176, 1402054, 9290842, 2114371, 19571216,
  10835491, 783926, 11785935, 4053824, 4233358, 12961683, 1095962, 5373555,
  919318, 7126489, 30503301, 3417735, 647464, 8715698, 7812880, 1770071,
  5910955, 584057)
)

library(dplyr)
library(stringr)

asthma_data <- asthma_data %>%
  mutate(
    prevalence = as.numeric(str_extract(prevalence_raw, "^[0-9.]+")),
    SE = as.numeric(str_extract(prevalence_raw, "(?<=\\().+?(?=\\))")),
    state_lower = tolower(state),
    region = state.region[match(state, state.name)]
  )

asthma_data <- na.omit(asthma_data)

```


# ABSTRACT


# BACKGROUND
Asthma is a chronic disease, presented as inflammation in the lungs and airways, that typically shows symptoms such as coughing, shortness of breath, tightness in the chest, and wheezing (Sockrider & Fussner, 2020). It also involves remodeling of the airway. This is seen in the thickening of the airway walls (Elias et. al, 1999). There are many different triggers that can lead to asthmatic symptoms.  These triggers include things such as strong odors, chemicals, dust, and allergens (Win and Hussain, 2009). Asthma, along with other allergic diseases has become more prevalent as urbanization increases (Kudo et. al, 2013). In many areas that are experiencing rapid urbanization and westernization, asthma prevalence has also increased (Paciência and Cavaleiro, 2020).  This paper focuses on the impact of higher population densities in urbanized areas on the asthma prevalence within that population.

```{r bar plot per state, fig.align='center', fig.width = 6, fig.height = 8}

library(ggplot2)

ggplot(asthma_data, aes(x = reorder(state, prevalence), y = prevalence)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Asthma Prevalence by U.S. State",
       x = "State",
       y = "Asthma Prevalence (%)") +
  theme_minimal()
```
This figure shows the Asthma Prevalence percentages per U.S. state.

# STUDY QUESTION and HYPOTHESIS

## Questions  
Does the rate of asthma increase in areas with higher population densities?

## Hypothesis 
Areas or states with a higher population density per square mile will have a higher prevalence of asthma.

## Prediction
Smaller states that have high populations will have a higher prevelance of asthma, and larger states with low populations will have lower prevelance of asthma.

# METHODS

```{r pop density graph, include=TRUE, eval=TRUE, warning=FALSE, fig.align='center', fig.width=12, fig.height=6}
# Load data
df <- read.csv("Comp Bio Data - Population Density.csv")
names(df)

# Load libraries
library(ggplot2)
library(viridis)  # Needed for viridis color scales

# Create plot
ggplot(df, aes(x = Population.Per.Square.Mile, y = Percent.With.Current.Athsma, colour = State)) +
  geom_point() +
 scale_colour_viridis_d() +
  theme_minimal() +
  labs(title = "Population Density vs Asthma Prevalence by State", x = 'Population Per Square Mile', y = 'Percent With Current Asthma') 

```
This plot shows the comparison by state of Population Density (per square mile) to Asthma Prevalence (Percent With Current Asthma)


```{r prevalence by state, include=TRUE, eval=TRUE, warning=FALSE, fig.align='center', fig.width=6, fig.height=3}

library(ggplot2)
library(maps)
library(dplyr)

# Map data
states_map <- map_data("state")

# Merge map data with asthma data
map_data <- states_map %>%
  left_join(asthma_data, by = c("region" = "state_lower"))

ggplot(map_data, aes(long, lat, group = group, fill = prevalence)) +
  geom_polygon(color = "gray90", size = 0.3) +  # state borders
  geom_polygon(aes(fill = prevalence), color = "white") +
  coord_fixed(1.3) +
 scale_fill_gradientn(
  name = "Asthma Prevalence (%)",
  colors = c("yellow2", "orange", "red", "red4")
) +
  labs(
    title = "Asthma Prevalence by U.S. State",
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12)
  )

```
This figure shows the prevalence of asthma by state with higher rates of asthma shown in darker colors.

```{r regions by color}
#just regions
# Get U.S. states map data
states_map <- map_data("state")

# Create a dataframe of states and their regions
state_regions <- data.frame(
  state = tolower(state.name),
  region = state.region
)

# Merge map data with region info
map_data <- states_map %>%
  left_join(state_regions, by = c("region" = "state"))

# Plot
ggplot(map_data, aes(x = long, y = lat, group = group, fill = region.y)) +
  geom_polygon(color = "white") +
  coord_fixed(1.3) +
  labs(title = "Four U.S. Census Regions",
       fill = "Region") +
  theme_void()
```



```{r scatter plot, include=TRUE, eval=TRUE, warning=FALSE, fig.align='center'}

ggplot(asthma_data, aes(x = population, y = prevalence)) +
  geom_point(aes(color=region)) +
  geom_text(aes(label = state_abrv), vjust = -0.5, size = 3)+
  labs(title = "Number of Individuals With Asthma vs. Asthma Prevalence by State",
       x = "Population",
       y = "Asthma Prevalence (%)") +
  theme_minimal()
```


```{r first analysis, eval=FALSE}

```



```{r second analysis, eval= FALSE}

```


# DISCUSSION


# CONCLUSION


# REFERENCES     
1. ChatGPT. OpenAI, version Jan 2025. Used as a reference for functions
    such as ggplot(), creating the map and troubleshooting error messages. Accessed 2025-11-10.
    
2. Centers for Disease Control and Prevention. (2024, November 21). Most recent asthma data. U.S. Department of Health and Human Services. https://www.cdc.gov/asthma-data/about/most-recent-asthma-data.html

3. Sockrider, M., & Fussner, L. (2020). What is asthma? | American Journal of Respiratory and Critical Care Medicine. https://www.atsjournals.org/doi/10.1164/rccm.2029P25 

4. Kudo, M., Ishigatsubo, Y., & Aoki, I. (2013, September 9). Pathology of asthma. Frontiers. https://www.frontiersin.org/journals/microbiology/articles/10.3389/fmicb.2013.00263/full 

5.https://pmc.ncbi.nlm.nih.gov/articles/PMC7152189/

6.https://www.jci.org/articles/view/8124?..&..&..&..

7.https://journals.lww.com/pbj/fulltext/2020/02000/Urban_level_environmental_factors_related_to.2.aspx